---
title: "Homework 4"
author: "PSTAT 115, Winter 2022"
date: "Due on February 27, 2022 at 11:59 pm"
output:
  pdf_document: default
  html_document:
    df_print: paged
urlcolor: blue
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo=TRUE,
                      cache=FALSE, 
                      fig.width=5, 
                      fig.height=5,
                      fig.align='center')
r = function(x, digits=2){ round(x, digits=digits) }
indent1 = '    '      
indent2 = paste(rep(indent1, 2), collapse='')
library(tidyverse)
library(reshape2)
library(magrittr)
library(testthat)
#install.packages("rstan")
library(rstan)
```


**Note:** If you are working with a partner, please submit only one
homework per group with both names and whether you are taking the course
for graduate credit or not. Submit your Rmarkdown (.Rmd) and the
compiled pdf on Gauchospace.

### Problem 1. Frequentist Coverage of The Bayesian Posterior Interval.

In the "random facts calibration game" we explored the importance and difficulty of well-calibrated
prior distributions by examining the calibration of subjective
intervals. Suppose that $y_1,..,y_n$ is an IID sample from a
$Normal(\mu, 1)$. We wish to estimate $\mu$.

**1a.** For Bayesian inference, we will assume the prior distribution
$\mu \sim Normal(0,\frac{1}{\kappa_0})$ for all parts below. Remember,
from lecture that we can interpret $\kappa_0$ as the pseudo-number of
prior observations with sample mean $\mu_0 = 0$. State the posterior
distribution of $\mu$ given $y_1,..,y_n$. Report the lower and upper
bounds of the $95\%$ quantile-based posterior credible interval for
$\mu$, using the fact that for a normal distribution with standard
eviation $\sigma$, approximately $95\%$ of the mass is between
$\pm 1.96\sigma$. \newline

 We know the Posterior Distribution is as follows:
$p(\theta|y,\sigma^2)=L(\mu)*p(\mu)$

Given $Y_i and \mu$ are normal, the posterior is also normally distributed
$p(\mu|y,\sigma)\propto L(\mu)*p(\mu)$

So, the posterior distribution can be justifies as $N(\mu_0,\tau^2_0)$. 

In our conjugacy model the parameters $\mu_0$ = $\frac{\frac{1}{\tau^2_0}\mu_0+\frac{n}{\sigma^2}\bar{y}}{\frac{1}{\tau^2_0}+\frac{n}{\sigma^2}}$, and $\tau^2_0$ = $\frac{1}{\frac{1}{\tau^2_0}+\frac{n}{\sigma^2}}$. 


In order to find the 95% confidence interval our interval we will need to plug in values for $\tau^2_0$ in terms of $K_0$ and $n$.

We are given a sampling variance of 1 and prior variance $1/k_0$. We also know our prior mean is 0.

We can now establish a lower and upper bounds for the model:
1.96*$\frac{1}{\frac{1}{\tau^2_0}+\frac{n}{\sigma^2}}$


Upper Bound: $1.96\sqrt{\frac{1}{n+k_0}}$
Lower Bound: $-1.96\sqrt{\frac{1}{n+k_0}}$
    

**1b**. Plot the length of the posterior credible interval as a function
of $\kappa_0$, for $\kappa_0 = 1, 2, ..., 25$ assuming $n=10$. Report
how this prior parameter effects the length of the posterior interval
and why this makes intuitive sense.

```{r}
k0 <- seq(1,25)
n = 10
values = c()
#Function
length_post_cred <- function(k0)
  {
    2*1.96*sqrt(1/(n+k0))
}

#Plot
interval_length <- ggplot(data.frame(k0),aes(k0)) + stat_function(fun=length_post_cred,geom= 'point', n = length(k0))
interval_length
```

```{r}
. = ottr::check("tests/q1b.R")
```
From the plot above we can conclude that as we get more samples and k0 increases, the variance for the estimate of MU decreases. This may suggest that our confidence in our estimate is increasing.


**1c**. Now we will evaluate the *frequentist coverage* of the posterior
credible interval on simulated data. Generate 1000 data sets where the
true value of $\mu=0$ and $n=10$. For each dataset, compute the
posterior $95\%$ interval endpoints (from the previous part) and see if
it the interval covers the true value of $\mu = 0$. Compute the
frequentist coverage as the fraction of these 1000 posterior 95%
credible intervals that contain $\mu=0$. Do this for each value of
$\kappa_0 = 1, 2, ..., 25$. Plot the coverage as a function of
$\kappa_0$.  Store these 25 coverage values in vector called `coverage`.

```{r}
## Fill in the vector called "coverage", which stores the fraction of intervals containing \mu = 0 for each kappa.
true_mu <- 0
n <- 10
data_count <- 1000
dataset <- matrix(0,data_count,length(k0))
coverage <- numeric(25)

for (k in k0) {
    for (data in seq(data_count)) {
        y <- rnorm(n,true_mu,1)
        post_mu <- (mean(y)*n/(k+n))
        cred_int <- qnorm(c(0.025,0.975),post_mu,sqrt(1/(k+n)))
        if(between(true_mu,cred_int[1],cred_int[2])==TRUE){
            dataset[data,k] <- 1
        }
    }
    coverage[k] <- sum(dataset[,k])/data_count
}
#plot
plot(k0,coverage,col="blue")
```

```{r}
. = ottr::check("tests/q1c.R")
```

**1d.** Repeat 1c but now generate data assuming the true $\mu=1$. Again, store these 25 coverage values in vector called `coverage`.

```{r}

## Fill in the vector called "coverage", which stores the fraction of intervals containing \mu = 1 for each kappa.
mu2 <- 1
data_count <- 1000
dataset <- matrix(0,data_count,length(k0))
coverage <- numeric(25)

for (k in k0) {
    for (data in seq(data_count)) {
        y <- rnorm(n,mu2,1)
        post_mu2 <- (mean(y)*n/(k+n))
        cred_int <- qnorm(c(0.025,0.975),post_mu2,sqrt(1/(k+n)))
        if(between(mu2,cred_int[1],cred_int[2])==TRUE){
            dataset[data,k] <- 1
        }
    }
    coverage[k] <- sum(dataset[,k])/data_count
}
#plot
plot(k0,coverage,col="red")
```

```{r}
. = ottr::check("tests/q1d.R")
```

**1e**. Explain the differences between the coverage plots when the true
$\mu$ = 0 and the true $\mu = 1$. For what values of $\kappa_0$ do you
see closer to nominal coverage (i.e. 95%)? For what values does your
posterior interval tend to over cover (the interval covers the true value
more than 95% of the time)? Undercover (the interval covers the true
value less than 95% of the time)? Why does this make sense?

When $\mu=0$, the posterior interval tends to over cover when k is greater than 2 and when k equals 1; It also tends to undercover when k is greater than 2.  we see that as $K_0$ increases, the nominal coverage increases for $\mu = 0$ but decreases for $\mu = 1$.


### Problem 2. Goal Scoring in the Women's World Cup

The Chinese Women's soccer team recently won the AFC Women's Asian Cup. Suppose you are interested in studying the World Cup performance of this soccer team. Let $\lambda$ be the be the average number of goals scored by the team. We will analyze $\lambda$ using the Gamma-Poisson model where data $Y_i$ is the observed number of goals scored in the $i$th World Cup game, ie. we have $Y_i|\lambda \sim Pois(\lambda)$.  _A priori_, we expect the rate of goal scoring to be $\lambda \sim Gamma(a, b)$. According to a sports analyst, they believe that $\lambda$ follows a Gamma distribution with $a = 1$ and $b = 0.25$. 

**2a.** 
Compute the theoretical posterior parameters a, b, and also the posterior mean. 

```{r}
library(rstan)
y <- c(4, 7, 3, 2, 3) # Number of goals in each game
n <- length(y)

a = 1
b = 0.25
post_a <- a + sum(y)
post_b <- b + length(y)
post_mu <- post_a/post_b
post_mu
```

```{r}
. = ottr::check("tests/q2a.R")
```

**2b.** 
Create a new Stan file by selecting "Stan file" and name it `women_cup.stan`, use Rstan to report and
estimate the posterior mean of the scoring rate by computing the sample
average of all Monte Carlo samples of $\lambda$.

```{r}
soccer_model <- stan_model("women_cup.stan")
stan_fit1 <- rstan::sampling(soccer_model,
                            data = list(N=n,y=y),
                            refresh = 0)
samples1 <- rstan::extract(stan_fit1)

lambda_samples <- samples1$lambda 

post_mean <- mean(lambda_samples) #help
post_mean
```

```{r}
. = ottr::check("tests/q2b.R")
```
**2c.**
Create a histogram of the Monte Carlo samples of $\lambda$ and add a line showing the theoretical posterior of density of $\lambda$. Do the Monte Carlo samples coincide with the theoretical density?

```{r}

soccer.df <- data.frame(
    lambda = as.numeric(lambda_samples)
    )

ggplot(soccer.df, aes(x=lambda)) + geom_histogram() + geom_vline(xintercept = post_mu, col = "pink")
```
Yes the Monte Carlo sample coincides with the theoretical density.


**2d.**
Use the Monte Carlo samples from Stan to compute the mean of predicative posterior distribution to estimate the distribution of expected goals scored for next game played by the Chinese women's soccer team.

```{r}
y <- c(4, 7, 3, 2, 3) # Number of goals in each game
n <- length(y)


#pred_mean <-
```

```{r}
#. = ottr::check("tests/q2d.R")
```

### Problem 3. Bayesian inference for the normal distribution in Stan.


Create a new Stan file and name it `IQ_model.stan`. We will make some basic modifications to the
template example in the default Stan file for this problem. Consider the
IQ example used from class. Scoring on IQ tests is designed to yield a
N(100, 15) distribution for the general population. We observe IQ scores
for a sample of $n$ individuals from a particular town,
$y_1, \ldots y_n \sim N(\mu, \sigma^2)$. Our goal is to estimate the
population mean in the town. Assume the
$p(\mu, \sigma) = p(\mu \mid \sigma)p(\sigma)$, where
$p(\mu \mid \sigma)$ is $N(\mu_0, \sigma/\sqrt{\kappa_0})$ and
$p(\sigma)$ is Gamma(a, b). Before you administer the IQ test you
believe the town is no different than the rest of the population, so you
assume a prior mean for $\mu$ of $\mu_0 = 100$, but you aren't to sure
about this a priori and so you set $\kappa_0 = 1$ (the effective number
of pseudo-observations). Similarly, a priori you assume $\sigma$ has a
mean of 15 (to match the intended standard deviation of the IQ test) and
so you decide on setting $a=15$ and $b=1$ (remember, the mean of a Gamma
is a/b). Assume the following IQ scores are observed:

```{r}
y <- c(70, 85, 111, 111, 115, 120, 123)
n <- length(y)
```

```{r}
#
```

**3a**. Make a scatter plot of the posterior distribution of the mean,
$\mu$, and the precision, $1/\sigma^2$. Put $\mu$ on the x-axis and
$1/\sigma^2$ on the y-axis. What is the posterior relationship between
$\mu$ and $1/\sigma^2$? Why does this make sense? *Hint:* review the
lecture notes. 

```{r}
normal_stan_model <- stan_model("IQ_model.stan")

k0 <- 1
mu0 <- 100
stan_fit2 <-rstan::sampling(normal_stan_model, data = list(N=n,y=y, mu0=mu0, k0=k0, a=a, b=b))
samples2 <-rstan::extract(stan_fit2)

mu_samples <- samples2$mu
sigma_samples <- samples2$sigma
precision_samples <- 1/((samples2$sigma)^2)

## Make the plot
tibble(Mean=mu_samples,Precision=1/sigma_samples^2) %>%
    ggplot()+geom_point(aes(x=Mean,y=Precision))
```

```{r}
. = ottr::check("tests/q3a.R")
```
Based on the data plot, we see that when there is a high precision, there is a low posterior variability is $\mu$. When there is a low precision, there is high uncertainty about $\mu$. This makes sense because $\sigma^2$ is a measure of spread.


**3b**. You are interested in whether the mean IQ in the town is greater
than the mean IQ in the overall population. Use Stan to find the
posterior probability that $\mu$ is greater than 100.

```{r}
y <- c(70, 85, 111, 111, 115, 120, 123)
n <- length(y)

mean(mu_samples>100)
```

The posterior probability of $\mu$ being greater than 100 is about 80%.

**3c.** You notice that two of the seven scores are significantly lower
than the other five. You think that the normal distribution may not be
the most appropriate model, in particular because you believe some
people in this town are likely have extreme low and extreme high scores.
One solution to this is to use a model that is more robust to these
kinds of outliers. The [Student's t
distribution](https://en.wikipedia.org/wiki/Student%27s_t-distribution)
and the [Laplace
distribution](https://en.wikipedia.org/wiki/Laplace_distribution) are
two so called "heavy-tailed distribution" which have higher
probabilities of outliers (i.e. observations further from the mean).
Heavy-tailed distributions are useful in modeling because they are more
robust to outliers. Fit the model assuming now that the IQ scores in the
town have a Laplace distribution, that is
$y_1, \ldots, y_n \sim Laplace(\mu, \sigma)$. Create a copy of the
previous stan file, and name it `IQ_laplace_model.stan1`. *Hint:* In the
Stan file you can replace `normal` with `double_exponential` in the
model section, another name for the Laplace distribution. Like the normal
distribution it has two arguments, $\mu$ and $\sigma$. Keep the same
prior distribution, $p(\mu, \sigma)$ as used in the normal model. Under
the Laplace model, what is the posterior probability that the median IQ
in the town is greater than 100? How does this compare to the
probability under the normal model? Why does this make sense?

```{r stan_samples, dependson="stan_def", cache=TRUE}

sm_t <- stan_model("IQ_laplace_model.stan")

k0 <- 1
mu0 <- 100
stan_fit3 <-rstan::sampling(sm_t, data = list(N=n,y=y, mu0=mu0, k0=k0, a=a, b=b), refresh=0)
samples3 <-rstan::extract(stan_fit3)

mu_samples <- samples3$mu
sigma_samples <- samples3$sigma

post_prob_laplace <- mean(mu_samples>100) #HELP
post_prob_laplace

```

```{r}
. = ottr::check("tests/q3c.R")
```
The posterior probability that the median IQ is greater than 100 is about 95%. This is much higher compared to the probability under the normal distribution. The biggest difference in the Laplace model is a conjugate model with the prior also following a normal distribution.